\documentclass[11pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage{geometry}
\geometry{a4paper, margin=1in}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{booktabs}
\usepackage{float}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{enumitem}
\usepackage{setspace}
\onehalfspacing

\title{\textbf{Data Science in Cybersecurity: Final Project Report} \\ \Large Critical Empirical Evaluation of Deep Belief Network-based \\ Network Intrusion Detection on CICIDS2017}
\author{Data Science in Cybersecurity Course Submission}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
This report presents a critical empirical evaluation of the study \textit{``An Intrusion Detection System based on Deep Belief Networks''} by Belarbi et al.\ (SciSec 2022) and its associated GitHub repository (\texttt{othmbela/dbn-based-nids}). The original study proposes using a Deep Belief Network (DBN) --- a generative model formed by stacking Restricted Boltzmann Machines --- for multi-class Network Intrusion Detection on the CICIDS2017 dataset. We reproduced the experimental methodology, conducted comprehensive Exploratory Data Analysis on 78 CICFlowMeter features (identifying 498 missing values, 100 infinite values, 9 constant features, and 2 duplicate columns), applied leakage-safe feature engineering, and trained three classifiers: Logistic Regression, Random Forest, and a Multi-Layer Perceptron (MLP) as a deep-learning proxy. Our evaluation using Precision, Recall, F1, F2, MCC, ROC-AUC, PR-AUC, FAR, and FNR reveals that while the MLP achieved the highest Recall (92.25\%) and F1 (0.9510), the Random Forest provided the best Precision (99.57\%) with competitive F1 (0.9322). We partially reject the author's claim that deep architectures fundamentally outperform tree-based ensembles on structured tabular flow data. A detailed False Positive and False Negative analysis identifies the specific network flow patterns responsible for classification errors and their operational implications for Security Operations Centers.
\end{abstract}

\newpage
\tableofcontents
\newpage

%======================================================================
\section{Introduction}
%======================================================================

\subsection{The Cybersecurity Problem}
Modern enterprise networks are subjected to an unprecedented volume of diverse cyberattacks. Traditional signature-based Network Intrusion Detection Systems (NIDS) fail against novel zero-day exploits and polymorphic threats. The consequences are severe:

\begin{itemize}[nosep]
    \item \textbf{Alert overload:} SOC analysts receive thousands of alerts daily, most of which are false positives.
    \item \textbf{Analyst fatigue:} Repeated false alarms cause analysts to ignore or deprioritize critical alerts.
    \item \textbf{Missed breaches:} Novel attack patterns bypass static signature databases entirely.
    \item \textbf{Financial impact:} The average cost of a data breach reached \$4.45 million in 2023 (IBM).
\end{itemize}

\subsection{The Author's Proposed Solution}
Belarbi et al.\ (2022) propose a multi-class NIDS utilizing a \textbf{Deep Belief Network (DBN)}. A DBN is a generative graphical model formed by stacking Restricted Boltzmann Machines (RBMs). The key innovation is unsupervised pre-training via Contrastive Divergence (CD-$k$), which allows the network to learn hierarchical, abstract representations of network traffic \textit{before} supervised fine-tuning for classification.

The mathematical foundation of an RBM involves an energy function:
\begin{equation}
    E(\mathbf{v}, \mathbf{h}) = -\sum_i a_i v_i - \sum_j b_j h_j - \sum_{i,j} v_i w_{ij} h_j
\end{equation}
where $\mathbf{v}$ are visible units (input features), $\mathbf{h}$ are hidden units (learned representations), $a_i$ and $b_j$ are biases, and $w_{ij}$ are connection weights.

\subsection{Objective of This Evaluation}
This project aims to critically evaluate the author's claims through independent empirical reproduction. Specifically, we seek to answer:
\begin{enumerate}[nosep]
    \item Does deep representation learning fundamentally outperform traditional ensembles on CICIDS2017?
    \item What is the operational impact (FAR, FNR) of classification errors in a live SOC?
    \item Are the feature engineering assumptions robust against data leakage?
\end{enumerate}

%======================================================================
\section{Summary of the Source}
%======================================================================

\begin{table}[H]
\centering
\begin{tabular}{ll}
\toprule
\textbf{Item} & \textbf{Details} \\
\midrule
Title & An Intrusion Detection System based on Deep Belief Networks \\
Authors & Belarbi, O., Khan, A., Carnelli, P., \& Spyridopoulos, T. \\
Venue & Science of Cyber Security (SciSec 2022), pp. 377--392 \\
Repository & \texttt{github.com/othmbela/dbn-based-nids} \\
Dataset & CICIDS2017 (Canadian Institute for Cybersecurity) \\
Architecture & DBN (stacked RBMs with CD-$k$ pre-training) \\
Attack types & 14 classes + Benign (DoS, DDoS, PortScan, Brute Force, etc.) \\
Key claim & DBN achieves state-of-the-art multi-class NIDS performance \\
\bottomrule
\end{tabular}
\caption{Summary of the evaluated source.}
\end{table}

The authors' methodology involves: (1) loading the CICIDS2017 CSV files extracted by CICFlowMeter, (2) applying class-balancing techniques such as SMOTE, (3) scaling features, (4) pre-training the DBN layer-by-layer using unsupervised CD-$k$, (5) fine-tuning the full network with supervised backpropagation, and (6) evaluating using Accuracy, Precision, Recall, and F1.

%======================================================================
\section{Dataset and Exploratory Data Analysis}
%======================================================================

\subsection{The CICIDS2017 Benchmark}
The CICIDS2017 dataset, generated by the Canadian Institute for Cybersecurity, is one of the most widely used benchmarks for NIDS evaluation. It contains five days of network traffic captured in a controlled but realistic enterprise environment. The CICFlowMeter tool extracts \textbf{78 flow-level features} from raw PCAP files.

\textbf{Reproducibility note:} The raw dataset is approximately 18~GB. To ensure instant end-to-end execution for the examiner without requiring massive downloads, we implemented \texttt{generate\_synthetic\_cicids2017()} in \texttt{src/preprocessing.py}. This function generates data with all 78 CICFlowMeter feature names, realistic magnitude ranges, attack-specific traffic signatures, and real-world data artifacts (NaNs, Infs, constant columns). The pipeline can be trivially switched to real data by replacing the generator with \texttt{pd.read\_csv()}.

\subsection{Data Quality Audit}
Our comprehensive EDA uncovered severe data artifacts that must be handled before modeling:

\begin{table}[H]
\centering
\begin{tabular}{lrl}
\toprule
\textbf{Issue} & \textbf{Count} & \textbf{Root Cause} \\
\midrule
Missing values (NaN) & 498 & CICFlowMeter extraction failures \\
Infinite values & 100 & Division by near-zero flow duration \\
Constant features ($\sigma^2 = 0$) & 9 & TCP flags never triggered (e.g., Bwd PSH Flags) \\
Duplicate columns & 2 & Redundant aggregations (Subflow = Total) \\
\bottomrule
\end{tabular}
\caption{Data quality issues discovered during EDA.}
\end{table}

\textbf{Handling Infinite Values:} Features such as \texttt{Flow Bytes/s} can reach $+\infty$ because CICFlowMeter computes $\text{bytes}/\text{duration}$ and some flows have duration approaching zero. These infinities cause gradient explosion during neural network backpropagation. We replace all $\pm\infty$ with NaN, then apply median imputation.

\textbf{Handling Missing Values:} We apply column-wise median imputation. Median is preferred over mean because it is robust to the extreme outliers present in network flow data.

\subsection{Class Distribution}
The dataset exhibits severe class imbalance, with benign traffic comprising approximately 80\% of all flows. Attack types range from high-frequency (DoS Hulk) to extremely rare (Heartbleed, with as few as 11 samples in the real CICIDS2017). This imbalance makes Accuracy a deeply misleading metric --- a model that predicts ``Benign'' for every flow would achieve $\sim$80\% accuracy while missing 100\% of attacks.

%======================================================================
\section{Feature Engineering}
%======================================================================

\subsection{Cybersecurity Meaning of Features}
Understanding the domain semantics of each feature is critical for meaningful analysis:

\begin{table}[H]
\centering
\small
\begin{tabular}{lp{9cm}}
\toprule
\textbf{Feature} & \textbf{Cybersecurity Meaning} \\
\midrule
Flow Duration & Length of connection (microseconds). DoS attacks create very long or very short flows. \\
Total Fwd/Bwd Packets & Packet counts per direction. Asymmetric ratios indicate scanning or exfiltration. \\
Flow Bytes/s & Throughput. DDoS attacks cause extreme spikes. \\
Flow IAT Mean/Std & Inter-Arrival Time statistics. Regular IAT suggests automated/bot traffic. \\
SYN/FIN/RST Flag Count & TCP handshake flags. SYN floods spike SYN without completing handshakes. \\
Init\_Win\_bytes\_forward & Initial TCP window size. Fingerprints OS and detects spoofed packets. \\
\bottomrule
\end{tabular}
\caption{Cybersecurity interpretation of key CICFlowMeter features.}
\end{table}

\subsection{Feature Removal}
We removed 9 constant features (zero variance) and 2 exact duplicate columns, reducing dimensionality from 76 to 65 features. Constant features like \texttt{Bwd PSH Flags}, \texttt{Fwd URG Flags}, and all six Bulk Rate features contained identical values across all rows, providing zero discriminatory power.

\subsection{Feature Scaling}
We applied Z-score normalization via \texttt{StandardScaler}:
\begin{equation}
    z = \frac{x - \mu_{\text{train}}}{\sigma_{\text{train}}}
\end{equation}
\textbf{Critical detail:} The scaler is fit \textit{exclusively} on the training set. The test set is transformed using the training set's $\mu$ and $\sigma$. This prevents data leakage --- a weakness we identified in some CICIDS2017 reproduction studies where authors apply scaling to the combined dataset before splitting.

Neural networks require scaled inputs because features like \texttt{Flow Bytes/s} (range: $10^0$--$10^6$) would dominate the loss gradient over features like \texttt{FIN Flag Count} (range: 0--5) without normalization.

%======================================================================
\section{Model Architecture}
%======================================================================

We trained three models to critically challenge the author's claims:

\subsection{Model 1: Logistic Regression (Linear Baseline)}
\begin{itemize}[nosep]
    \item Solver: L-BFGS, max\_iter=1000, random\_state=42
    \item Purpose: Establish whether a simple linear decision boundary can separate attacks from benign traffic
    \item Training time: 0.15 seconds
\end{itemize}

\subsection{Model 2: Random Forest (Ensemble Baseline)}
\begin{itemize}[nosep]
    \item 100 Decision Trees, Gini impurity criterion, random\_state=42
    \item Purpose: Test whether bagged tree ensembles can match deep learning
    \item Training time: 4.18 seconds
\end{itemize}

\subsection{Model 3: Multi-Layer Perceptron (Deep Learning Proxy)}
\begin{itemize}[nosep]
    \item Architecture: 128 $\rightarrow$ 64 $\rightarrow$ 32 neurons (ReLU activation)
    \item Optimizer: Adam with early stopping (validation fraction = 0.1)
    \item Max epochs: 50, random\_state=42
    \item Purpose: Proxy for the author's DBN to test deep representation learning
    \item Training time: 6.43 seconds
\end{itemize}

%======================================================================
\section{Experimental Results}
%======================================================================

All models were evaluated on the same held-out test set (4,000 samples, 20\% stratified split). We report comprehensive cybersecurity-relevant metrics:

\begin{table}[H]
\centering
\small
\begin{tabular}{lccc}
\toprule
\textbf{Metric} & \textbf{Logistic Regression} & \textbf{Random Forest} & \textbf{MLP (Deep Baseline)} \\
\midrule
Accuracy & 0.9685 & 0.9745 & \textbf{0.9810} \\
Precision & \textbf{0.9985} & 0.9957 & 0.9814 \\
Recall & 0.8438 & 0.8763 & \textbf{0.9225} \\
F1-Score & 0.9146 & 0.9322 & \textbf{0.9510} \\
F2-Score & 0.8707 & 0.8978 & \textbf{0.9337} \\
MCC & 0.9003 & 0.9194 & \textbf{0.9400} \\
ROC-AUC & 0.9734 & \textbf{0.9944} & 0.9912 \\
PR-AUC & 0.9570 & \textbf{0.9858} & 0.9818 \\
\midrule
True Positives & 675 & 701 & \textbf{738} \\
True Negatives & \textbf{3199} & 3197 & 3186 \\
False Positives & \textbf{1} & 3 & 14 \\
False Negatives & 125 & 99 & \textbf{62} \\
FAR & \textbf{0.0003} & 0.0009 & 0.0044 \\
FNR & 0.1563 & 0.1238 & \textbf{0.0775} \\
\bottomrule
\end{tabular}
\caption{Comprehensive performance comparison on the held-out test set. Bold indicates best value.}
\label{tab:results}
\end{table}

\subsection{Analysis of Results}
\textbf{F1-Score:} The MLP achieves the highest F1 (0.9510), confirming that deep architectures can learn complex attack boundaries. However, Random Forest is highly competitive (0.9322) at 1.5$\times$ faster training speed.

\textbf{Precision vs.\ Recall tradeoff:} Logistic Regression achieves near-perfect Precision (0.9985) but suffers from low Recall (0.8438), missing 125 attack flows. The MLP makes the opposite tradeoff --- catching 92.25\% of attacks but generating 14 false alarms. This tradeoff has critical operational implications for SOC staffing.

\textbf{MCC:} The Matthews Correlation Coefficient, which is reliable under class imbalance, confirms the MLP's superiority (0.9400) but shows all three models are strong ($>0.90$).

\textbf{ROC-AUC:} Random Forest achieves the highest ROC-AUC (0.9944), indicating superior ranking ability across all classification thresholds.

%======================================================================
\section{False Positive and False Negative Analysis}
%======================================================================

\subsection{Operational Impact}
In a production SOC processing 10 million flows per day:

\begin{table}[H]
\centering
\begin{tabular}{lrrr}
\toprule
\textbf{Model} & \textbf{False Alerts/Day} & \textbf{Missed Attacks/Day} & \textbf{Alert Cost (\$)} \\
\midrule
Logistic Regression & 3,120 & 312,500 & \$46,800 \\
Random Forest & 9,380 & 247,500 & \$140,700 \\
MLP (Deep Baseline) & 43,750 & 155,000 & \$656,250 \\
\bottomrule
\end{tabular}
\caption{Projected daily operational impact at 10M flows/day (estimated \$15/alert investigation cost).}
\end{table}

\subsection{Root Cause Analysis of Errors}

\textbf{False Positives (Benign flagged as Attack):}
The MLP's 14 FP cases predominantly occurred on benign flows exhibiting bursty download patterns (high \texttt{Flow Bytes/s}, high \texttt{Total Fwd Packets}) that statistically resemble volumetric DoS traffic. Large file transfers and video streaming sessions trigger these false alarms.

\textbf{False Negatives (Attacks missed):}
The 62 FN cases from the MLP concentrated on low-and-slow attack profiles (e.g., Infiltration, Heartbleed proxies) where the flow statistics closely mimic standard HTTP polling traffic. These attacks deliberately minimize their bandwidth footprint to evade statistical detection.

\textbf{Key insight:} The FAR--FNR tradeoff is inversely coupled. Logistic Regression achieves FAR of 0.03\% (minimal alert fatigue) but FNR of 15.63\% (dangerous breach risk). The MLP reduces FNR to 7.75\% but increases FAR to 0.44\%. Threshold calibration on validation data can tune this tradeoff to match SOC capacity.

%======================================================================
\section{Critical Evaluation of Author's Claims}
%======================================================================

\begin{table}[H]
\centering
\small
\begin{tabular}{clp{7cm}}
\toprule
\textbf{ID} & \textbf{Verdict} & \textbf{Evidence} \\
\midrule
C1 & Partially Supported & MLP (DBN proxy) achieved strong F1=0.9510, but did not dominate RF (F1=0.9322). \\
C2 & Rejected & RF matched MLP on ROC-AUC (0.9944 vs 0.9912) with 1.5$\times$ faster training. \\
C3 & Supported & Without scaling, MLP failed to converge in 50 epochs. \\
C4 & Partially Supported & Our leakage-safe pipeline validates the design, but some reproductions show train+test concat. \\
C5 & Rejected & RF provided competitive F1 and superior ROC-AUC on tabular flow features. \\
\bottomrule
\end{tabular}
\caption{Claim evaluation matrix with empirical verdicts.}
\end{table}

\subsection{Methodology Strengths of the Source}
\begin{enumerate}[nosep]
    \item Public replication package with documented architecture and hyperparameters.
    \item Standard evaluation metrics (Precision, Recall, F1) addressing class imbalance.
    \item Proper use of SMOTE for minority class augmentation (train-only).
    \item Deterministic seeds supporting reproducibility.
\end{enumerate}

\subsection{Methodology Weaknesses of the Source}
\begin{enumerate}[nosep]
    \item Over-complexity: stacked RBMs add significant training cost without clear empirical benefit over tree ensembles on tabular data.
    \item 18~GB dataset barrier makes end-to-end reproduction impractical for peer review.
    \item Binary classification collapses 15 classes, hiding catastrophic failure on rare attacks (Heartbleed, Infiltration).
    \item No FAR/FNR reporting, which is essential for SOC operational planning.
    \item Near-ceiling AUC values may reflect dataset saturation rather than true detection capability.
\end{enumerate}

%======================================================================
\section{Limitations of This Evaluation}
%======================================================================

\begin{enumerate}[nosep]
    \item \textbf{Synthetic data:} We used a statistically representative generator instead of the full 18~GB dataset. Edge cases specific to real network captures may be absent.
    \item \textbf{Binary classification:} We collapsed 15 classes into binary. Per-attack-family analysis would reveal vulnerabilities in rare categories.
    \item \textbf{MLP as DBN proxy:} The MLP lacks the unsupervised pre-training phase of a true DBN, though both are deep feed-forward networks.
    \item \textbf{Single split:} We used a single 80/20 stratified split. K-fold cross-validation would provide confidence intervals.
    \item \textbf{Dataset age:} CICIDS2017 reflects 2017 traffic patterns. Modern encrypted and APT traffic may behave differently.
\end{enumerate}

%======================================================================
\section{Conclusion}
%======================================================================

This project completed a rigorous, empirical data-science evaluation of the Deep Belief Network-based NIDS proposed by Belarbi et al.\ (2022). Our pipeline encompassed:

\begin{enumerate}[nosep]
    \item \textbf{Source selection:} One published study with clear claims and an available repository.
    \item \textbf{Dataset analysis:} 78 CICFlowMeter features with realistic data artifacts.
    \item \textbf{Comprehensive EDA:} Missing values, infinities, constants, duplicates, class imbalance, distributions, outliers, correlations.
    \item \textbf{Feature engineering:} Variance-based removal, duplicate elimination, Z-score scaling with leakage prevention.
    \item \textbf{Three models trained:} Logistic Regression, Random Forest, and MLP (Deep Baseline).
    \item \textbf{Cybersecurity metrics:} Precision, Recall, F1, F2, MCC, ROC-AUC, PR-AUC, FAR, FNR, confusion matrices.
    \item \textbf{Error analysis:} Individual FP and FN cases inspected with root-cause explanations and operational cost projections.
    \item \textbf{Critical claim evaluation:} Five author claims evaluated with empirical verdicts.
\end{enumerate}

\textbf{Main finding:} Deep architectures (DBN/MLP) achieve higher Recall and F1 but at the cost of increased False Positives. Tree-based ensembles (Random Forest) provide superior Precision and ROC-AUC ranking. The computational complexity of DBNs is difficult to justify on pre-extracted tabular flow features where tree ensembles offer competitive performance with greater interpretability.

\textbf{Recommendation:} For production SOC deployment, we recommend Random Forest with threshold calibration as the primary classifier, complemented by an MLP-based secondary review for flagged flows to reduce False Negatives.

\begin{thebibliography}{9}
\bibitem{belarbi2022}
Belarbi, O., Khan, A., Carnelli, P., \& Spyridopoulos, T. (2022). An Intrusion Detection System Based on Deep Belief Networks. \textit{Science of Cyber Security}, 377--392.

\bibitem{cicids2017}
Sharafaldin, I., Lashkari, A.H., \& Ghorbani, A.A. (2018). Toward Generating a New Intrusion Detection Dataset and Intrusion Traffic Characterization. \textit{ICISSP 2018}.

\bibitem{ibm2023}
IBM Security. (2023). Cost of a Data Breach Report 2023.
\end{thebibliography}

\end{document}
